---
title: "Per-task extraction economics: a token & cost benchmark across the optimization rollout"
author: "nlmixr2 project"
date: "2026-06-25"
output: rmarkdown::html_document
---

<!--
Reproducible benchmark companion to ingestion-challenges.Rmd (the methods-paper
draft). The per-task data are a frozen snapshot baked into
inst/extdata/token_optimization_2026-06-25.csv so this article re-knits without
the live runner. Refresh by re-exporting that CSV from the runner state files
and bumping the date.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
f <- system.file("extdata", "token_optimization_2026-06-25.csv", package = "nlmixr2libingest")
if (!nzchar(f)) f <- "../../inst/extdata/token_optimization_2026-06-25.csv"
d <- read.csv(f, stringsAsFactors = FALSE)
has_kable <- requireNamespace("knitr", quietly = TRUE)
```

## Purpose

This article measures the **per-task token and dollar cost** of literature
model extraction, split by outcome (a model was **built** vs the paper was a
**no-extraction** skip), and compares a **recent** cohort against an older
**baseline** cohort. The motivating question: *how much benefit are the token
optimizations delivering?*

The cost model (see `ingestion-challenges.Rmd`) is dominated by
`cache_read x agent-turns`, so we report the cost drivers — cache-read and
output tokens — alongside total tokens and the runner's dollar estimate.

## Data and method

- **Source.** Per-task aggregates summed across all runs of each `completed`
  task, read from the runner state files (frozen snapshot).
- **Cohorts.**
  - `recent_24h` — completed in the 24 h before the snapshot
    (2026-06-25 11:28 UTC). This is **after** the agent-side *policy
    self-check* (merged 2026-06-23) and its *strengthening* (2026-06-24).
  - `baseline_48h+` — the 50 most-recent tasks completed **more than 48 h**
    before the snapshot that carry a deliverable report (so they are
    classifiable). This is **before** the policy self-check.
  - `preopt_pre0622` — **all** tasks completed **before 2026-06-22** (n ≈ 1565),
    i.e. **before Phase 0** (register-lookup + R-log-filter). These predate the
    deliverable-report convention, so they cannot be split extraction-vs-skip;
    they are reported **overall (dollar + tokens) only**. Cost is reliable here —
    **100%** of these records carry a populated `cost_usd` (vs gaps in the
    report-era cohorts) — and the cohort is **extraction-dominated** (≈76% have
    >12k output tokens).
- **Outcome classifier.** Report-based: a task is a `no_extraction` skip if its
  `reports/<task>.md` deliverable contains skip language (e.g. "NOT a fittable
  model", "(skip)", "methodology paper", "conference poster ... not"); otherwise
  `extraction`.

```{r cohorts}
tab <- as.data.frame.matrix(table(d$cohort, d$cls))
tab$total <- rowSums(tab)
if (has_kable) knitr::kable(tab, caption = "Task counts by cohort and outcome")
```

> **Read the confounds before the numbers.** Both cohorts are **post-Phase-0**:
> the register-lookup (`lookup_canonical`) and R-log-filter optimizations landed
> 2026-06-22, so *both* cohorts already have them. The cohort difference is
> therefore mainly the **policy self-check**, which targets *sidecar frequency*,
> not per-task tokens. To reach **before** Phase 0 we add the `preopt_pre0622`
> cohort — but only at the **dollar/overall** level (it cannot be split
> extraction-vs-skip), and its task *composition* differs from the report-era
> cohorts, so cross-era comparison is confounded (see the dollar-baseline
> section).

## Results

```{r helpers}
summ1 <- function(x) {
  x <- x[is.finite(x)]
  c(n = length(x), mean = mean(x), sd = if (length(x) > 1) sd(x) else 0,
    median = median(x), p25 = unname(quantile(x, .25)), p75 = unname(quantile(x, .75)))
}
fmt <- function(v, money) {
  if (money) sprintf("$%.2f", v)
  else ifelse(abs(v) >= 1000, sprintf("%.0fk", v / 1000), sprintf("%.0f", v))
}
metrics <- c(cost = "cost (USD)", total = "total tokens",
             cache_read = "cache-read tokens", output = "output tokens")
build <- function(cls) {
  rows <- list()
  for (m in names(metrics)) for (co in c("recent_24h", "baseline_48h+")) {
    s <- summ1(d[d$cls == cls & d$cohort == co, m])
    money <- m == "cost"
    rows[[length(rows) + 1]] <- data.frame(
      metric = metrics[m],
      cohort = ifelse(co == "recent_24h", "recent (post-policy)", "baseline (>48h)"),
      n = s["n"],
      mean = fmt(s["mean"], money), sd = fmt(s["sd"], money),
      median = fmt(s["median"], money), p25 = fmt(s["p25"], money), p75 = fmt(s["p75"], money),
      row.names = NULL, check.names = FALSE)
  }
  do.call(rbind, rows)
}
```

### Extraction (a model was built)

```{r extraction}
if (has_kable) knitr::kable(build("extraction"),
  caption = "Extraction tasks — mean / sd / median / p25 / p75 (small n; medians unstable)")
```

### No-extraction (skip)

```{r noextraction}
if (has_kable) knitr::kable(build("no_extraction"),
  caption = "No-extraction (skip) tasks — mean / sd / median / p25 / p75")
```

### Pre-Phase-0 dollar baseline (before 2026-06-22)

These tasks predate the report convention, so they are reported **overall** (no
extraction/skip split). The estimated dollar cost is the requested baseline.

```{r preopt}
ov <- function(co, m) {
  x <- d[d$cohort == co, m]; x <- x[is.finite(x)]
  c(n = length(x), mean = mean(x), sd = if (length(x) > 1) sd(x) else 0,
    median = median(x), p25 = unname(quantile(x, .25)), p75 = unname(quantile(x, .75)))
}
eras <- c(preopt_pre0622 = "pre-Phase-0 (<06-22)",
          `baseline_48h+` = "baseline (06-22..23)",
          recent_24h = "recent (post-policy)")
ctab <- do.call(rbind, lapply(names(eras), function(co) {
  c1 <- ov(co, "cost"); t1 <- ov(co, "total")
  data.frame(era = eras[co], n = c1["n"],
             mean = sprintf("$%.2f", c1["mean"]), sd = sprintf("$%.2f", c1["sd"]),
             median = sprintf("$%.2f", c1["median"]), p25 = sprintf("$%.2f", c1["p25"]),
             p75 = sprintf("$%.2f", c1["p75"]),
             `tok median` = sprintf("%.0fk", t1["median"] / 1000),
             row.names = NULL, check.names = FALSE)
}))
if (has_kable) knitr::kable(ctab,
  caption = "Overall per-task COST (USD) & token median by era — UNCLASSIFIED (mixes extraction + skip; composition differs across eras)")
```

> **Composition warning — do not read the overall drop as savings.** The eras
> differ in task *mix*: pre-Phase-0 is extraction-dominated (older popPK papers),
> while the two report-era cohorts are skip-dominated (the Metrum/vendor index
> batch). So the overall fall from a ~\$3 median to ~\$0.8 is mostly the cheaper
> skip mix, **not** an optimization effect. The only within-class read available
> — pre-Phase-0 overall (≈extraction-dominated, median ~\$3.2) vs the report-era
> *extraction* medians (\$7.5 recent / \$15 baseline) — runs the *other* way; but
> those post-extractions are more complex Metrum models (TMDD, survival,
> multi-analyte), so model difficulty confounds that read too. Net: **no clean
> dollar evidence either way** — the cross-era signal is composition, not tooling.

## Controlling for model complexity (archetype-matched)

The dollar gap above is partly *what kind of model* each era extracted. To remove
that, we classify every extraction's model into a structural archetype
(compartment count, elimination type, PD/TMDD/TTE structure) and compare cost
**within** archetype across eras. Pre-era extractions are read from the merged
registry; post-era extractions are isolated from their task worktrees (the new
`inst/modeldb` file vs `origin/main`).

> The `nlmixr2libingest` classifier (`build_feature_table()` /
> `assign_archetype()`) is the intended tool, but currently errors against the
> installed `nlmixr2lib` (`modeldb` not exported), so a deterministic structural
> parse stands in here. Fixing the package classifier is tracked separately.

```{r archetypes}
fa <- system.file("extdata", "token_optimization_archetypes_2026-06-25.csv", package = "nlmixr2libingest")
if (!nzchar(fa)) fa <- "../../inst/extdata/token_optimization_archetypes_2026-06-25.csv"
a <- read.csv(fa, stringsAsFactors = FALSE)
a$era <- factor(a$era, levels = c("pre0622", "baseline", "recent"),
                labels = c("pre-Phase-0", "baseline", "recent"))
agg <- do.call(rbind, lapply(split(a, list(a$archetype, a$era), drop = TRUE), function(g) {
  data.frame(archetype = g$archetype[1], era = as.character(g$era[1]), n = nrow(g),
             median = sprintf("$%.2f", median(g$cost)),
             p25 = sprintf("$%.2f", quantile(g$cost, .25)),
             p75 = sprintf("$%.2f", quantile(g$cost, .75)), row.names = NULL)
}))
agg <- agg[order(agg$archetype, agg$era), ]
if (has_kable) knitr::kable(agg, row.names = FALSE,
  caption = "Per-task cost (USD) by model archetype and era — extraction tasks only")
```

**What it shows.**

- **Complexity dominates cost.** In the pre-Phase-0 corpus, simple linear PK runs
  ~\$9–10 median, while TMDD, indirect-response / Emax PD and TTE/survival run
  \$13–17 — a 1.5–2× premium. That quantifies the difficulty confound behind the
  overall numbers.
- **Like-for-like, no per-task saving.** The cleanest comparison —
  **simple linear PK**, well-populated in every era — is **flat**: ~\$9.3
  (pre-Phase-0) vs ~\$8.6–9.5 (recent). Pre-Phase-0 here is genuinely *before* the
  register-lookup / R-log-filter tools, so this is the Phase-0-inclusive
  comparison the overall section could not make — and it is flat.
- **Complex archetypes read *more* expensive post**, not less (TMDD ~\$20 vs ~\$8;
  TTE ~\$25 vs ~\$14), but each post cell holds only 1–6 tasks and the post
  (Metrum-index) instances are atypically hard *within* an archetype (e.g.
  meta-analytic survival) — noise plus residual difficulty, not a regression.

**Bottom line:** complexity-matching confirms the dollar verdict — no per-task
cost reduction is visible from the token optimizations, even against a true
pre-Phase-0 baseline. The realized win remains sidecar reduction.

## Interpretation

1. **Skips — the bulk of both cohorts (`r sum(d$cls=="no_extraction")` of
   `r nrow(d)` tasks) — are flat.** Median cost is ~\$0.7–0.8 and median total
   tokens ~0.5 M in both cohorts (recent slightly *higher*). This is the
   expected result: the skip path (read the paper → decide not-fittable → write
   a note) is **not a target** of the per-task token optimizations, and the
   policy self-check changes *whether a sidecar is filed*, not the cost of
   reaching a skip decision.

2. **Extractions are too small-n to conclude (`r sum(d$cls=="extraction")`
   tasks total).** Mean cost is modestly lower in the recent cohort, but the
   token medians move the other way — a signature of small-sample instability,
   not a real effect. The baseline extraction sample (n≈9) is also atypically
   light. No reliable per-task token *reduction* is demonstrated here.

3. **The dominant driver — cache-read tokens — is essentially unchanged**
   between cohorts. The cached per-turn context has not shrunk measurably, which
   is consistent with point 1: the policy work did not touch what the agent
   carries per turn.

4. **The realized benefit of the recent work is elsewhere: sidecar reduction.**
   Across the recent cohort the new-sidecar rate is ~1–2% of completions (vs a
   large historical backlog), i.e. far fewer stop/restart cycles. That is an
   operator-time and throughput win, **not** a per-task-token win — and this
   benchmark, which holds per-task tokens fixed, is exactly where you would
   *not* see it.

## Limitations

- **Small extraction n** (recent ≈14, baseline ≈9); medians swing on individual
  tasks. Treat the extraction rows as directional at best.
- **`cost_usd` is partially under-populated** in the run records (cost and token
  medians land on different tasks), so **token counts are the firmer metric**;
  dollar figures are indicative.
- **Cohorts are Metrum/vendor-index-dominated**, which is heavily skip-weighted —
  few extraction samples to show the build-loop optimizations in.
- **Report-based classifier**; the baseline is restricted to report-bearing
  (post-2026-06-22) tasks, so it cannot reach pre-Phase-0 work.
- **Both cohorts are post-Phase-0**, so this isolates the *policy* work, not the
  register-lookup / R-log-filter optimizations.

## Conclusion

In this window the policy optimizations did **not** reduce per-task tokens — and
they were not expected to; their value is fewer sidecars. Reaching back to a
**pre-Phase-0 dollar baseline** (median ≈\$3.2/task overall, extraction-dominated)
does not change the verdict: the cross-era cost signal is dominated by task
*composition* (extraction-heavy → skip-heavy) and *difficulty* (post-era Metrum
models are more complex), not by tooling, so there is **no clean dollar evidence
of per-task savings** in either direction. **This snapshot is therefore best used
as the baseline** against which future Phase 1–3 work
(archetype templates, distillation, budget escalation) should be A/B-measured —
ideally with a per-task `POLICY_SELFCHECK` / optimization marker so the
extraction-vs-skip split and the optimization cohort are recorded at the source
rather than reconstructed.
